Precise Data Flow Analysis\\in the Presence of Correlated Method Calls

### 2015-03-09 sas2015/content/concl.tex
squishing (mainly) of introduction

Previous algorithms for data-flow analysis are unable to avoid propagating
data-flow facts along infeasible paths that arise in the presence of
correlated polymorphic method calls. 
% 
We present an approach for transforming an IFDS problem into an IDE problem
in which path feasibility is encoded into the functions that are associated
with edges in an exploded control-flow supergraph. The solution to this
IDE problem can be mapped back to the solution space of the original IFDS
problem, and is more precise for some client programs because the flow of data along
infeasible paths is prevented. We present a formalization of the
transformation and prove its correctness. We also briefly report on
preliminary experiments with our prototype implementation. Full proof
details are available as supplemental materials with this submission,
and we plan to submit our implementation and experimental data to the
artifact evaluation committee.

>>>>

Previous algorithms for data-flow analysis are unable to avoid propagating
data-flow facts along infeasible paths that arise in the presence of
correlated polymorphic method calls. 
% 
We present an approach for transforming an IFDS problem into an IDE problem
in which path feasibility is encoded into the functions that are associated
with edges in an exploded control-flow supergraph. The solution to this
IDE problem can be mapped back to the solution space of the original IFDS
problem, and is more precise for some client programs because the flow of data along
infeasible paths is prevented. We present a formalization of the
transformation and prove its correctness. We also briefly report on
preliminary experiments with our prototype implementation. Full proof
details are available in the Appendix,
and we plan to submit our implementation and experimental data to the
artifact evaluation committee.

### 2015-03-09 sas2015/content/example.tex
applying ondrej's corrections
squishing (mainly) of introduction

\subsection{Occurrences of Correlated Calls}\vspace{-.5mm}
How often do correlated calls occur in practice? To assess the benefit of the correlated-calls analysis, we counted the number of correlated calls that occur in programs of the Dacapo benchmarks~\cite{blackburn2006dacapo}, using the WALA framework for static analysis on Java bytecode~\cite{fink2012wala}.
Our goal was to obtain an upper bound on the number of redundant IFDS-result nodes that could be potentially removed by our analysis. The results are shown in Table~\ref{tab:dacapostat} in the Appendix.

>>>>

\subsection{Occurrences of Correlated Calls}\label{sec:occur}\vspace{-.5mm}
How often do correlated calls occur in practice? To assess the benefit of the correlated-calls analysis, we counted the number of correlated calls that occur in programs of the Dacapo benchmarks~\cite{blackburn2006dacapo}, using the WALA framework~\cite{fink2012wala}.
Our goal was to obtain an upper bound on the number of redundant IFDS-result nodes that could be potentially removed by our analysis. The results are shown in Table~\ref{tab:dacapostat} in the Appendix.

### 2015-03-09 sas2015/content/impl.tex
squishing (mainly) of introduction

To compare two micro-functions for equality, it suffices to compare the corresponding
sets $I(r)$ and $U(r)$ for all receivers $r$. The following lemma shows that this
implementation of comparison corresponds to equality of the represented micro-functions:
\begin{restatable}{lemma}{equalEq}
\label{lem:equalEq}
    For any pair of micro-function representations $\setpair{I}{U}$, $\setpair{I'}{U'}$,
    \begin{equation}
        \forall r\,.\,I(r) = I'(r) \land U(r) = U'(r)
        \iff
        \denote{\setpair{I}{U}} = \denote{\setpair{I'}{U'}}.
    \end{equation}
\end{restatable}
The proofs of all of the lemmas and theorems in the paper are provided in the supplemental
material submitted with the paper.

@@

\subsection{Implementation of the Correlated-Calls Analysis}
The correlated-calls analysis was implemented in the Scala programming language~\cite{odersky2004overview}. 
The input programs that we analyzed were written in Java. Our implementation analyzed the
corresponding Java bytecode.
To retrieve information about an input program, such as its control-flow supergraph or the set of receivers and their types, we used the WALA framework for static analysis~\cite{fink2012wala}.
Since WALA currently only contains an implementation of IFDS, we implemented IDE from scratch. Instead of using WALA's IFDS implementation, to run an IFDS problem, we converted it to an IDE problem and used our own IDE solver.

>>>>

To compare two micro-functions for equality, it suffices to compare the corresponding
sets $I(r)$ and $U(r)$ for all receivers $r$. The following lemma shows that this
implementation of comparison corresponds to equality of the represented micro-functions:
\begin{restatable}{lemma}{equalEq}
\label{lem:equalEq}
    For any pair of micro-function representations $\setpair{I}{U}$, $\setpair{I'}{U'}$,
    \begin{equation}
        \forall r\,.\,I(r) = I'(r) \land U(r) = U'(r)
        \iff
        \denote{\setpair{I}{U}} = \denote{\setpair{I'}{U'}}.
    \end{equation}
\end{restatable}
The proofs of all of the lemmas and theorems are in the Appendix.

@@

\subsection{Implementation of the Correlated-Calls Analysis}
The correlated-calls analysis was implemented in the Scala programming language~\cite{odersky2004overview}. 
The input programs that we analyzed were written in Java. Our implementation analyzed the
corresponding Java bytecode.
To retrieve information about an input program, such as its control-flow supergraph or the set of receivers and their types, we used the WALA framework~\cite{fink2012wala}.
Since WALA currently only contains an implementation of IFDS, we implemented IDE from scratch. Instead of using WALA's IFDS implementation, to run an IFDS problem, we converted it to an IDE problem and used our own IDE solver.

### 2015-03-09 sas2015/content/intro.tex
tweaks
squishing (mainly) of introduction

Data-flow analysis computes an approximation of how values may flow  
through a program, and has applications in compiler optimization, 
programming tools, and computer security, and many other areas.
Data-flow analyses operate on \textit{control-flow graphs} (CFGs)
that model 
the order in which the instructions of a program are executed. This is
typically done by associating \textit{flow functions} that represent how 
data is propagated with the edges of the CFG.  A 
\textit{confluence operator} specifies how the data facts that have been 
computed along different paths should be merged when the paths join.

Since a CFG is an over-approximation of the possible flows 
of control in concrete executions of a program, it may contain \textit{infeasible} 
paths that cannot occur at runtime. The precision of a data-flow analysis algorithm
depends critically on its ability to detect and disregard such infeasible paths.
The popular \textit{Interprocedural Finite Distributive Subset} (IFDS) algorithm 
by  Reps, Horwitz, and Sagiv~\cite{reps1995precise} is a general data-flow analysis
algorithm for computing solutions to standard finite distributive data-flow problems such 
as reaching definitions, available expressions, and taint analysis.
A distinguishing 
characteristic of IFDS is that it avoids infeasible interprocedural paths in which calls 
and returns to/from functions are not properly matched. Sagiv, Reps, and Horwitz
also presented the \textit{Interprocedural Distributive Environment} (IDE) algorithm
\cite{sagiv1996precise} that similarly only considers properly matched call/return
edges, but that supports a broader range of data-flow problems by expanding the 
domain of flow functions to \textit{environments} that go beyond the data-flow facts 
considered by IFDS.

This paper presents an approach to data-flow analysis that avoids a type
of infeasible path that arises in object-oriented programs when two or more
methods are dynamically dispatched on the same receiver object.
%\mtodo{Should we define the term ``receiver''?}
In such cases, if the
method calls are polymorphic (i.e., if they dispatch to different method definitions 
depending on the type of the receiver expression at run time), then their dispatch 
behaviours will be correlated. A recent paper~\cite{DBLP:journals/scp/Tip15} identified
this problem but did not present a concrete solution or algorithm, and we are not aware 
of any existing data-flow analysis that is capable of avoiding infeasible paths that arise 
in the presence of correlated method calls.
% (see Section~\ref{sec:Related} for a discussion
%of related work).

@@

The approach taken in our work is to transform a standard IFDS problem into an
IDE problem that precisely accounts for infeasible paths due to correlated calls. 
The results of this IDE problem can be mapped back to the data-flow domain of the 
original IFDS problem.

We present a formalization of the transformation and prove its correctness
as follows. 
First, we derive an IDE problem that is equivalent to the original IFDS problem,
as described by Reps et al.~\cite{sagiv1996precise}.
Solving this ``Equivalence-IDE Problem'' with the IDE algorithm yields a result that
is equivalent to the one obtained by applying the IFDS algorithm to the original IFDS problem.
Then, we derive a ``Correlated-Calls 
IDE Problem'' from the original IFDS problem and show 
that the solution to this problem, when converted to an IFDS result, is more precise
than the original IFDS result.
We also show that the correlated-calls analysis is sound, i.e., that it never considers concrete execution paths as infeasible.
%This is illustrated schematically
%in Figure~\ref{fig:transformations}%
%\footnote{ 
%  The labels %$\resultIDE$, $\resultIFDS$, $\backEq$, $\backCC$, $\transEq$, and $\transCC$
%  on edges in 
%  Figure~\ref{fig:transformations} reflect a number of mappings and %projections 
%  that will be defined in Section~\ref{sec:cca} and that can be %ignored here.
%}.

@@

The remainder of this paper is organized as follows.
%
Section~\ref{sec:MotivatingExample} presents a motivating example.
%
Section~\ref{sec:bg} reviews the IFDS and IDE algorithms.
%
Section~\ref{sec:cca} presents the correlated-calls transformation
and sketches a proof of its correctness%
\footnote{
  Detailed proofs of our lemmas and theorems can be found in
  the supplemental materials that are included with this submission.
}.
%
Related work is discussed in Section~\ref{sec:Related}.
%
Finally, conclusions and directions for future work are presented in Section~\ref{sec:Conclusions}.

>>>>

\commentout{
Data-flow analysis computes an approximation of how values may flow  
through a program, and has applications in compiler optimization, 
programming tools, and computer security, and many other areas.
Data-flow analyses operate on \textit{control-flow graphs} (CFGs)
that model 
the order in which the instructions of a program are executed. This is
typically done by associating \textit{flow functions} that represent how 
data is propagated with the edges of the CFG.  A 
\textit{confluence operator} specifies how the data facts that have been 
computed along different paths should be merged when the paths join.
}

%Since a CFG is an over-approximation of the possible flows 
A control-flow graph (CFG) is an over-approximation of the possible flows 
of control in concrete executions of a program. It may contain \textit{infeasible} 
paths that cannot occur at runtime. The precision of a data-flow analysis algorithm
depends on its ability to detect and disregard such infeasible paths.
The \textit{Interprocedural Finite Distributive Subset} (IFDS) algorithm~\cite{reps1995precise}
is one general data-flow analysis algorithm that 
%by  Reps, Horwitz, and Sagiv
%is a general data-flow analysis
%algorithm
%for computing solutions to finite distributive data-flow problems such 
%that can be instantiated to solve problems such
%as reaching definitions, available expressions, and taint analysis.
%A distinguishing 
%characteristic of IFDS is that it
avoids infeasible interprocedural paths in which calls 
and returns to/from functions are not properly matched. %Sagiv, Reps, and Horwitz
%also presented
The \textit{Interprocedural Distributive Environment} (IDE) algorithm~\cite{sagiv1996precise}
has the same property, but
%that similarly only considers properly matched call/return
%edges, but that 
supports a broader range of data-flow problems.
%by expanding the 
%domain of flow functions to \textit{environments} that go beyond the data-flow facts 
%considered by IFDS.

This paper presents an approach to data-flow analysis that avoids a type
of infeasible path that arises in object-oriented programs when two or more
methods are dynamically dispatched on the same receiver object.
%\mtodo{Should we define the term ``receiver''?}
%In such cases,
If the
method calls are polymorphic
(i.e. the method invoked depends on the run-time type of the receiver),
%(i.e., if they dispatch to different method definitions 
%depending on the type of the receiver expression at run time),
then their dispatch 
behaviours are correlated, and some of the paths between them are infeasible. A recent paper~\cite{DBLP:journals/scp/Tip15} made this observation
but did not present any concrete algorithm to take advantage of it.
%, and we are not aware 
%of any existing data-flow analysis that is capable of avoiding infeasible paths that arise 
%in the presence of correlated method calls.
% (see Section~\ref{sec:Related} for a discussion
%of related work).

@@

%The approach taken in our work is to 
Our approach transforms an IFDS problem into an
IDE problem that precisely accounts for infeasible paths due to correlated calls. 
The results of this IDE problem can be mapped back to the data-flow domain of the 
original IFDS problem, but are more precise than the results of directly applying
the IFDS algorithm to the original problem.
%
We present a formalization of the transformation and prove its correctness:
specifically, we prove it still soundly considers all paths that are feasible,
and that it avoids flow along all paths that are infeasible due to
correlated calls.
%as follows. 
%First, we derive an IDE problem that is equivalent to the original IFDS problem,
%as described by Reps et al.~\cite{sagiv1996precise}.
%Solving this ``Equivalence-IDE Problem'' with the IDE algorithm yields a result that
%is equivalent to the one obtained by applying the IFDS algorithm to the original IFDS problem.
%Then, we derive a ``Correlated-Calls 
%IDE Problem'' from the original IFDS problem and show 
%that the solution to this problem, when converted to an IFDS result, is more precise
%than the original IFDS result.
%We also show that the correlated-calls analysis is sound, i.e., that it never considers concrete execution paths as infeasible.
%This is illustrated schematically
%in Figure~\ref{fig:transformations}%
%\footnote{ 
%  The labels %$\resultIDE$, $\resultIFDS$, $\backEq$, $\backCC$, $\transEq$, and $\transCC$
%  on edges in 
%  Figure~\ref{fig:transformations} reflect a number of mappings and %projections 
%  that will be defined in Section~\ref{sec:cca} and that can be %ignored here.
%}.

@@

The remainder of this paper is organized as follows.
%
Section~\ref{sec:MotivatingExample} presents a motivating example.
%
Section~\ref{sec:bg} reviews the IFDS and IDE algorithms.
%
Section~\ref{sec:cca} presents the correlated-calls transformation
and sketches a proof of its correctness%
\footnote{
  Detailed proofs of our lemmas and theorems can be found in
  the appendix.
}.
%
Related work is discussed in Section~\ref{sec:Related}.
%
Finally, conclusions and directions for future work are presented in Section~\ref{sec:Conclusions}.
